Modelling the emergence of whisker barrels

### 2020-01-27 paper/briefpaper/brief.tex
Adds methods section covering computation of divJ

% Blue for comments
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}

@@

\cmnt{It's still a short report if I add a condensed version of supp.tex
  here.}

\cmnt{Software for modelling the cortical sheet as a hexagonal lattice with
  no-flux boundary conditions and an arbitrary boundary shape, for computing
  the flux using a finite difference method, for numerical integration by the
  fourth-order Runge-Kutta method, and for reproducing these results is
  available at \url{https://github.com/ABRG-Models/BarrelEmerge}}

>>>>

% Blue for comments
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\newcommand{\code}[1]{\textsf{#1}}

@@

\cmnt{I've left this section in black, although it is all new, cf the PNAS
  draft.}
% Be sure to include/mention:
%
% The no-flux boundary conditions applied in the \cite{karbowski_model_2004}
% model are retained.
%
% Arbitrary boundary shapes were applied
%
% computing the flux using a finite difference method,
%
% numerical integration by the fourth-order Runge-Kutta method,

In these simulations, the cortex was modelled as a two dimensional lattice in
which each element represented a small area of the cortical sheet. We chose to
use hexagonal elements to simplify the computation of the 2D Laplacian, a
spatial operator used in reaction diffusion systems. Within a boundary traced
from rat primary barrel cortex (Fig.\,1A) we set the hex-to-hex distance
$d=0.03$\,mm, which resulted in a lattice containing 6515 hexes for the
simulations shown in Figs.\,1A,C \& D and 12739 hexes for the Fgf8
misexpression study shown in Fig.\,1E. Each hex contained 82 time-dependent
variables: 41 branching densities ($a_i$) and 41 connection densities
($c_i$). Each hex also contained 82 time-independent variables; the $x$ and
$y$ components of the sum of the product of the interaction paramters with the
two molecular signalling gradients.

The rates of change of each of the time-dependent variables (Eqs.\,\ref{eq:dc}
\& \ref{eq:da}) was computed using a fourth-order Runge-Kutta method. The most
involved part of this computation was finding the divergence of the flux of
axonal branching, $\mb{J}_i(\mb{x},t)$---the term in parentheses in
Eq.\,\ref{eq:da}:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) \equiv \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right)
\end{equation}
%
We can define the time-independent sum of the guidance gradients
$\mb{g}_i(\mb{x}) \equiv \sum_{j=1}^{M} \gamma_{i,j} \nabla \rho_j(\mb{x})$
and then apply vector calculus identities to compute this term. Because
the divergence operator is distributive, Eq.~\ref{eq:divJ} can be expanded
(dropping all explicit references to space or time dependence):
%
\begin{equation}
\nabla\vcdot\mb{J}_i = \nabla\vcdot\big(D \nabla a_i\big) - \nabla\vcdot\big(a_i \mb{g}_i\big)
\end{equation}
%
Because $D$ is a constant and by applying the vector calculus product rule
identity to $\nabla\vcdot\big(a_i \mb{g}_i\big)$ we obtain:
%
%%\begin{equation}
%%\nabla\vcdot\mb{J}_i = D \dvrg a_i
%%-
%%\big(
%%a_i\nabla\vcdot\mb{g}_i
%%+
%%\mb{g}_i\vcdot\nabla a_i
%%\big)
%%\end{equation}
%
%%(to prove $\nabla\vcdot D\nabla a = D \dvrg a$, apply the
%%product rule and note that $\nabla D = 0$). Expanding
%%the brackets we have:
%
\begin{equation} \label{eq:divJExpanded}
%
\nabla\vcdot\mb{J}_i =
%
D \dvrg a_i % term1
-
a_i\nabla\vcdot\mb{g}_i % term2
-
\mb{g}_i\vcdot\nabla a_i % term3
%
\end{equation}
%
which has three elements to compute: i)~$D \dvrg a_i$ (the Laplacian of
$a_i(\mb{x},t)$); ii)~a time-independent modulator of $a_i(\mb{x},t)$ (because
$\nabla\vcdot\mb{g}_i(\mb{x})$ is a time-independent static field); and iii)~the
scalar product of the static vector field $\mb{g}_i$ and the gradient of
$a_i$.

Each of the divergences can be simplified by means of Gauss's Theorem
following \cite{lee_hexagonal_2014}. The computation of the mean value of the
Laplacian across the area of one hexagon of area $\Omega$ located at position
$\mb{p}_0$, with neighbours at positions $\mb{p}_1$--$\mb{p}_6$ is:
%
%%\begin{equation} \label{eq:computeDivJTerm1}
%%\boxed{
%%D \dvrg a_i(\mb{p}_0,t) = D\frac{2}{3d^2} \sum_{j=1}^{6}\big(a_i(\mb{p}_j) - a_i(\mb{p}_0)\big)
%%}
%%\end{equation}

\begin{equation}
\begin{split}
\langle D \dvrg a_i(\mb{p}_0,t) \rangle  = \frac{1}{\Omega} \iint_A \dvrg a_i(\mb{x}, t) dxdy & = \frac{1}{\Omega} \oint_c \frac{\partial a_i}{\partial \hat{\mb{n}}} d\gamma \\
& \approx \frac{1}{\Omega} \sum_{j=1}^6 \frac{\partial a_i(\mb{p}_j)}{\partial \hat{\mb{n}}} \bigg\rvert_{mid} v \\
& = \frac{2}{\sqrt{3} d^2} \sum_{j=1}^6 \frac{a_i(\mb{p}_j) - a_i(\mb{p}_0)}{d} \frac{d}{\sqrt{3}} \\
& = \frac{2}{3 d^2} \sum_{j=1}^6 \big(a_i(\mb{p}_j) - a_i(\mb{p}_0)\big)
\end{split}
\end{equation}

where $d$ is the centre-to-centre distance between hexes in the grid, $v =
d/\sqrt{3}$ is the length of each of the 6 edges of the hexagon's perimeter
and $d\gamma$ is an infinitessimally small distance along the perimeter of the
hexagon. The area, $\Omega$, of each hexagon in the lattice is
$\frac{\sqrt{3}}{2}d^2$.

The computation of the second term in Eq.\,\ref{eq:divJExpanded},
$\langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}(\mb{p}_0)\rangle$, can be written out similarly:

\begin{equation} \label{eq:divg}
\begin{split}
%
\frac{1}{\Omega} \iint_A a_i\nabla\vcdot\mb{g}\;dxdy & = \frac{a_i(\mb{p}_0)}{\Omega}  \oint_c \mb{g}\vcdot d\hat{\mathbf{n}} \\
%
& \approx \frac{a_i(\mb{p}_0,t)}{\Omega} \sum_{j=1}^{6} \frac{\mb{g}_j + \mb{g}_0}{2}\vcdot \hat{\mb{n}}\;v \\
%
& = \frac{2 a_i(\mb{p}_0,t)v}{\sqrt{3}d^2} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
%%& = \frac{2 a_i(\mb{p}_0,t)d}{\sqrt{3}d^2\sqrt{3}} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
\Rightarrow \langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}(\mb{p}_0)\rangle & = \frac{a_i(\mb{p}_0,t)}{3d} \bigg( \sum_{j=1}^{6} \big(\mb{g}_j^x + \mb{g}_0^x\big) \cos (60\times(j-1)) + \sum_{j=1}^{6} \big({\mb{g}_j^y + \mb{g}_0^y}\big) \sin (60\times(j-1)) \bigg) \\
%
\end{split}
\end{equation}

The final expression is suitable for a numerical computation and can
be found in \code{compute\_divJ()} as \code{term2}.

The final term in Eq.~\ref{eq:divJExpanded} is the scalar product of two
vector fields which is straighforward to compute from their Cartesian
components.

The advantage of separating this computation out is that we can ensure that,
at the boundary, $\mb{J}$ resulting from the first term remains 0, by the
ghost cell method, then $\mb{g}_i(\mb{x})$ can be tailored so that it, and its
normal derivative approach 0 at the boundary, ensuring that the second and
third terms of Eq.~\ref{eq:divJExpanded} also contribute nothing to
$\mb{J}$. This is achieved by applying a fairly sharp logistic function based
on distance to the boundary to $\mb{g}(\mb{x})$.

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge}
